### 2019-12-09 ic-journal/contraction.tex
Up to Section 4.0

Milner's ``unique solution of equations'' theorem for $\wb$
(Theorem~\ref{t:Mil89})
brings a new proof technique for proving (weak) bisimilarities. However, it has
 limitations: the equations must be guarded and sequential (moreover
all sums where equation variables appear must be guarded sums).
This limits the usefulness of the technique, since
the occurrences of other operators above the variables, such as parallel
composition and restriction, in general cannot be removed.
The constraints of Theorem~\ref{t:Mil89}, however, can be 
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
where $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
$Q$ \emph{has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
Process $Q$, however, may be non-deterministic and may have other ways
of doing the same work, and these could be slow (i.e., involving
more $\tau$-actions than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

\begin{definition}[contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \emph{(bisimulation) contraction} if, whenever
 $P\RR Q$, %for all $\mu$

\begin{enumerate}
\item   $P \arr\mu P'$ implies there is $Q'$ such that $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$   implies there is $P'$ such that $P \Arcap \mu
 P'$ and $P' \wb Q'$\enspace.
\end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
contraction}, written as $P \mcontrBIS Q$ ($P$ \emph{contracts to} $Q$),
if $P\ \R\ Q$ for some contraction $\R$.
Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.
\end{definition}

In the first clause $Q$ is required to match $P$'s challenge
transition with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, the second clause of Definition~\ref{d:BisCon}, on the
challenges from $Q$, entirely ignores efficiency: it is the same
clause of  weak bisimulation~--- the final derivatives are indeed required
to be related  by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than 
the \emph{expansion relation}
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}:
\begin{definition}[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever
we have $P\RR Q$, for all $\mu$
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$   implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
  expansion}, written as $P \expa Q$ ($P$ \emph{expands} $Q$),
 if $P \RR Q$ for some expansion $\R$.
 \end{definition}
The expansion preorder is widely used in proof techniques for bisimilarity and
intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the both preorders, while in clause (2) expansion uses 
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$; 
moreover in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.

@@

In the same way as bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar, see e.g.~\cite{sangiorgi2017equations}.

>>>>

Milner's ``unique solution of equations'' theorem for $\wb$
(Theorem~\ref{t:Mil89})
brings a new proof technique for proving (weak) bisimilarities. However, it has
 limitations: the equations must be guarded and sequential. (\hl{Moreover,}
all sums where equation variables appear must be guarded sums.)
This limits the usefulness of the technique, since
the occurrences of other operators \hl{using equation} variables, such as parallel
composition and restriction,
in general cannot be \hl{eliminated}. % without changing the meaning of equations
The constraints \hl{in} Theorem~\ref{t:Mil89}, however, can be
weakened if we move from equations to a special kind of inequations called
  \emph{contractions}.

Intuitively, the bisimilarity contraction $\mcontrBIS$ is a preorder
\hl{that} $P \mcontrBIS Q$ holds if $P \wb Q$ and, in addition, 
\emph{$Q$ has the possibility of being at least as efficient as $P$} (as far as
$\tau$-actions are performed).
The process $Q$, however, may be nondeterministic and may have other ways
\hl{to do} the same work, \hl{ways} which could be \hl{slower} (i.e., involving
more $\tau$-actions than those performed by $P$).
% Thus, in contrast with expansion,  we cannot really say that `$Q$ is more efficient than
% $P$'.

\begin{definition}[contraction]
\label{d:BisCon}
A process relation ${\R}$ 
 is a \emph{(bisimulation) contraction} if, whenever $P\RR Q$,

\begin{enumerate}
\item $P \arr\mu P'$ implies \hl{that} there is $Q'$ \hl{with} $Q \arcap \mu
  Q'$ and $P' \RR Q'$;
\item $Q \arr\mu Q'$ implies \hl{that} there is $P'$ \hl{with} $P \Arcap \mu
 P'$ and $P' \wb Q'$.
\end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
contraction}, written as $P \mcontrBIS Q$,
if $P\ \R\ Q$ for some contraction $\R$.
Sometimes we write $\mexpaBIS$ for the inverse of $\mcontrBIS$.
\end{definition}
In clause (1) \hl{of the above definition}, $Q$ is required to match the challenge
transition \hl{of $P$} with at most one transition.
This makes sure that $Q$ is capable of mimicking % verb: mimic
$P$'s work at least as efficiently as $P$. 
In contrast, \hl{clause (2) entirely ignores efficiency on the challenges from $Q$:}
the final derivatives are required to be related by $\wb$, rather than by $\R$.

Bisimilarity contraction is coarser than bisimilarity expansion
$\expa$~\cite{arun1992efficiency,sangiorgi2015equations}, \hl{one of the
most useful auxiliary relations in up-to techniques}:
\begin{definition}[expansion]
\label{d:expa}
A process relation ${\R}$
  is an \emph{expansion} if, whenever $P\RR Q$,
 \begin{enumerate}
 \item   $P \arr\mu P'$ implies that there is $Q'$ with $Q \arcap \mu  Q'$
  and $P' \RR Q'$;
 \item $Q \arr\mu Q'$ implies that there is $P'$ with $P \Arr \mu P'$ and $P' \RR Q'$.
 \end{enumerate}
Two processes $P$ and $Q$ are in the \emph{bisimilarity
  expansion}, written as $P \expa Q$, if $P \RR Q$ for some expansion $\R$.
 \end{definition}
\hl{Bisimilarity expansion} is widely used in proof techniques for bisimilarity.
\hl{It} intuitively refines bisimilarity by 
formalising the idea of ``efficiency'' between processes.
Clause (1) is the same in the both preorders, while in clause (2) expansion \hl{requires}
$P \Arr \mu P'$, rather than $P \Arcap \mu P'$.
\hl{Moreover,} in clause (2) of Def.~\ref{d:BisCon} the final derivatives
are simply required to be bisimilar ($P' \wb Q'$).
Intuitively, $P \expa Q$ holds if $P\wb Q$ and, in addition, \emph{$Q$
  is always at least as efficient as $P$}.

@@

\hl{Bisimilarity expansion and bisimilarity contraction are both
preorders. (Thus we can also call them the \emph{expansion or contraction
preorder}, respectively.)}
Similarily with (weak) bisimilarity, both the expansion and the
contraction preorders are preserved by all CCS operators except the
summation. The proofs are similar \hl{with the case of bisimilarity},
see, e.g.~\cite{sangiorgi2017equations} \hl{for details.}

### 2019-12-09 ic-journal/formal.tex
Up to Section 4.0

We highlight here a formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to LCF
\cite{gordon1979edinburgh,milner1972logic} by Robin Milner and others
since 1972. It is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in the HOL family (e.g. Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed HOL has considerably simpler logical
foundations than most other theorem provers. %, e.g. Coq. 
As a consequence, theories and proofs verified in HOL are easier to understand
 to people who are not familar with more advanced
dependent type theories.

>>>>

We highlight here a \hl{comprehensive} formalisation of CCS in the HOL theorem prover
(HOL4)~\cite{Melham:1993vl,slind2008brief}, with a focus towards the
theory (and formal proofs) of the unique solution of
equations/contractions theorems mentioned in Section~\ref{s:eq} and
\ref{s:mcontr}.
All proof scripts are available as part of HOL's official
examples\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL/tree/master/examples/CCS}}.
The work so far consists of about 24,000 lines (1MB) of code in total,
in which about 5,000 lines were derived from the early work of Monica
Nesi~\cite{Nesi:1992ve} on HOL88, with major modifications.

Higher Order Logic (HOL)~\cite{hollogic} traces
its roots back to the
\emph{Logic of Computable Functions
  (LCF)}~\cite{gordon1979edinburgh,milner1972logic}
by Robin Milner and others
since 1972. It is a variant of
Church’s Simple Theory of Types (STT)~\cite{church1940formulation},
plus a higher order version of Hilbert's choice operator $\varepsilon$,
Axiom of Infinity, and Rank-1 (prenex) polymorphism.
HOL4 has implemented the original HOL, 
while some other theorem provers in the HOL family (e.g.~Isabelle/HOL) have
certain extensions.
%  (they made the formal language more powerful,
% but they also bring the possibilities that the entire logic becomes
% inconsistent). 
Indeed, HOL has considerably simpler logical
foundations than most other theorem provers.
As a consequence, theories and proofs verified in HOL are easier to understand
 \hl{for} people who are not familar with more advanced
dependent type theories, \hl{e.g.~those implemented by Coq.}

### 2019-12-09 ic-journal/unique.tex
Up to Section 4.0

A \emph{system of contractions} is defined as a system of equations,
except that the contraction symbol $\mcontr$ is used in the place of
$=$ in Def.~\ref{def:equation}. Thus a system of contractions is a set 
$\{  X_i \mcontr E_i\}_{i\in I}$ where $I$ is an indexing set and each
$E_i$  contains variables in $\til X$.

@@

\begin{lemma}
\label{l:uptocon}
Suppose $\til P$ and $\til Q$ are solutions for $\mcontrBIS$
 of a system of weakly-guarded contractions that uses 
guarded sums.
For any context $\qct$  that uses guarded sums,
if $\ct{\til P}\Arr{\mu}  R$, then there is a context $\qctp$  that uses 
guarded sums
such that $R \mcontrBIS \ctp{\til P}$ and $\ct{\til Q} \Arcap{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}

\begin{proof}{(sketch from~\cite{sangiorgi2017equations})}
  Let $n$ be the length (the number of one-step transitions) of
a transition $\ct{\til P}\Arr\mu R$, and
let $\ctpp {\til P}$ and $\ctpp {\til Q}$  be the processes obtained
from  $\ct {\til P}$ and $\ct {\til Q}$ by unfolding the definitions
of the contractions $n$ times. Thus in $\qctpp$ each hole is
underneath at least $n$ prefixes, and cannot contribute to an action
in the first $n$ transitions; moreover all the contexts use guarded sums.

We have $\ct{\til P} \mcontrBIS \ctpp{\til P}$, and 
$\ct{\til Q} \mcontrBIS \ctpp{\til Q}$, 
 from the precongruence property of $\mcontrBIS$ (we exploit here
 the syntactic constraints on sums). Moreover,
 since each hole of the  context $\qctpp$ is underneath at least $n$
 prefixes, applying the definition
 of $ \mcontrBIS$ on the transition 
 $\ct{\til P}\Arr{\mu}  R$, we infer the existence
 of $\qctp$ such that 
$\ctpp{\til P}\Arcap{\mu} \ctp{\til P} \mexpaBIS R$
and 
$\ctpp{\til Q}\Arcap{\mu}  \ctp{\til Q}$.
Finally, again applying the definition of $\mcontrBIS$ on 
$\ct{\til Q} \mcontrBIS \ctpp{\til Q}$, 
we derive 
$\ct{\til Q}\Arcap{\mu} R' \wb \ctp{\til Q}$ for some $R'$.
\end{proof}

@@

Theorem~\ref{t:contraBisimulationU} brings a new proof technique for
bisimilarity, which is  less restrictive than Milner's
Theorem~\ref{t:Mil89} (but with the additional costs of checking $\mcontrBIS$
in addition to $\wb$).
However, comparing with Milner's Theorem~\ref{t:Mil89s1}, there remains
a constraint: the involved CCS expression may only
use guarded sums. This is mainly due to the fact that $\wb$ is not a
 congruence and also $\mcontrBIS$ is not a  precongruence.
Inspired by rooted bisimilarity, to  eliminate the restriction
on guarded sums we refine the idea of contractions by moving
to \emph{rooted contractions}:

@@

The above definition was found with the help of interactive theorem proving.
The following two principles were adopted when searching
for a possible definition: (1) the definition should not be coinductive,
along the lines of rooted bisimilarity $\rapprox$ (Def.~\ref{d:rootedBisimilarity});
(2) the definition should be built on top of  the existing \emph{contraction}
relation $\mcontrBIS$.
Furthermore, we needed to show that the definition \hl{being found}
indeed yields the coarsest precongruence
\hl{contained} in $\mcontrBIS$. The proof is similar with
the analogous result for $\rapprox$. See Section~\ref{s:coarsest} for
more details.

@@

For this new relation, the analogous of Lemma~\ref{l:uptocon} and of
Theorem~\ref{t:contraBisimulationU} can now be stated without
constraints on the sum operator.
The schema of the proofs is almost identical, because all 
properties of $\rcontr$ needed in this proof is its precongruence, which is
indeed true on unrestricted contexts including direct sums:
\begin{lemma}
\label{l:ruptocon}
Suppose $\til P$ and $\til Q$ are solutions  for $\rcontr$ 
 of a system of weakly-guarded contractions.
For any context $\qct$, 
if  $\ct{\til P}\Arr{\mu}  R$,
 then there is a context $\qctp$
such that $R \mcontrBIS \ctp{\til P}$ and  $\ct{\til Q} \Arr{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}

@@

\begin{theorem}[Unique solution of rooted contractions]
\label{t:rcontraBisimulationU}
Let $E_i$ be weakly guarded with free variables in $\til X$,
and let $\til P \rcontr \til E\{\til P/\til X\}$,
    $\til Q \rcontr \til E\{\til Q/\til X\}$. Then $\til P \rapprox \til Q$.
\end{theorem}

>>>>

A \emph{system of contractions} is defined as a system of equations,
except that the contraction symbol $\mcontrBIS$ is used in the place of
$=$ in Def.~\ref{def:equation}. Thus a system of contractions is a set 
$\{  X_i \mcontrBIS E_i\}_{i\in I}$ where $I$ is an indexing set and each
$E_i$  contains variables in $\til X$.

@@

\begin{lemma}
\label{l:uptocon}
Suppose $\til P$ and $\til Q$ are solutions (for $\mcontrBIS$)
 of a system of weakly guarded contractions that uses guarded sums.
For any context $\qct$  that uses guarded sums,
if $\ct{\til P}\Arr{\mu}  R$, then there is a context $\qctp$ that uses 
guarded sums
such that $R \mcontrBIS \ctp{\til P}$ and $\ct{\til Q} \Arcap{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}

\begin{proof}{(sketch from~\cite{sangiorgi2017equations})}
  Let $n$ be the length (i.e., the number of one-step transitions) of
a transition $\ct{\til P}\Arr\mu R$, and
let $\ctpp {\til P}$ and $\ctpp {\til Q}$  be the processes obtained
from  $\ct {\til P}$ and $\ct {\til Q}$ by unfolding the \hl{definition
of $\mcontrBIS$ for} $n$ times. Thus in $\qctpp$ each hole is
underneath at least $n$ prefixes, and \hl{therefore} cannot contribute to an action
in the first $n$ transitions. \hl{Moreover}, all \hl{involved} contexts use guarded sums.

We have $\ct{\til P} \mcontrBIS \ctpp{\til P}$ and 
$\ct{\til Q} \mcontrBIS \ctpp{\til Q}$
 from the precongruence property of $\mcontrBIS$ (we exploit here
 the syntactic constraints on sums). Moreover,
 since each hole of the  context $\qctpp$ is underneath at least $n$
 prefixes, applying the definition
 of $\mcontrBIS$ on the transition 
 $\ct{\til P}\Arr{\mu}  R$, we infer the existence
 of $\qctp$ such that 
$\ctpp{\til P}\Arcap{\mu} \ctp{\til P} \mexpaBIS R$
and 
$\ctpp{\til Q}\Arcap{\mu}  \ctp{\til Q}$.
Finally, again applying the definition of $\mcontrBIS$ on 
$\ct{\til Q} \mcontrBIS \ctpp{\til Q}$, 
we derive 
$\ct{\til Q}\Arcap{\mu} R' \wb \ctp{\til Q}$ for some $R'$.
\end{proof}

@@

Theorem~\ref{t:contraBisimulationU} brings a new proof technique for
bisimilarity, which is  less restrictive than Milner's
Theorem~\ref{t:Mil89} (but with the additional costs of checking $\mcontrBIS$
in addition to $\wb$).
However, comparing with Milner's Theorem~\ref{t:Mil89s1}, there remains
\hl{one limitation}: the involved CCS expression \hl{can} only
use guarded sums. This is mainly due to the fact that $\wb$ is not a
 congruence and also $\mcontrBIS$ is not a precongruence.
Inspired by rooted bisimilarity, to eliminate the restriction
on guarded sums we refine the idea of contractions by moving
to \emph{rooted contractions}:

@@

The above definition was found with the help of interactive theorem proving.
The following two principles were adopted when \hl{manually} searching
for a possible definition: (1) the definition should not be coinductive,
along the lines of rooted bisimilarity $\rapprox$ (Def.~\ref{d:rootedBisimilarity});
(2) the definition should be built on top of  the existing \emph{contraction}
relation $\mcontrBIS$.
Furthermore, we needed to \hl{prove} that the definition \hl{being found}
indeed yields the coarsest precongruence
\hl{contained} in $\mcontrBIS$. The proof is similar with
the analogous result for $\rapprox$. See Section~\ref{s:coarsest} for
more details.

@@

For this new relation, the analogous of Lemma~\ref{l:uptocon} and of
Theorem~\ref{t:contraBisimulationU} can now be stated without
constraints on \hl{summation}.
The schema of the proofs is almost identical, because all 
properties of $\rcontr$ needed in this proof is its precongruence, which is
indeed true for \hl{all weakly guarded contexts:} % removed "direct sums"
\begin{lemma}
\label{l:ruptocon}
Let $\til P$ and $\til Q$ be solutions (for $\rcontr$)
 of a system of weakly guarded contractions.
For any context $\qct$, 
if  $\ct{\til P}\Arr{\mu}  R$,
 then there is a context $\qctp$
such that $R \mcontrBIS \ctp{\til P}$ and  $\ct{\til Q} \Arr{\mu} R'
 \wb \ctp{\til Q}$ for some $R'$.
\end{lemma}

@@

\begin{theorem}[unique solution of rooted contractions]
\label{t:rcontraBisimulationU}
Let $E_i$ be weakly guarded with free variables \hl{up to} $\til X$,
and let $\til P \rcontr \til E\{\til P/\til X\}$,
    $\til Q \rcontr \til E\{\til Q/\til X\}$. Then $\til P \rapprox \til Q$.
\end{theorem}
